### 2025-04-15 multiple-regression.qmd
20250415 - mediation

>>>>

A mediation model takes the following general form in the `lavaan` package [@Rosseel2012; @R-lavaan].

mediationModel <- '
  # direct effect (cPrime)
  Y ~ direct*X

  # mediator
  M ~ a*X
  Y ~ b*M

  # indirect effect = a*b
  indirect := a*b

  # total effect (c)
  total := direct + indirect
'
```

Let's substitute in our [predictor](#sec-correlationalStudy), [outcome](#sec-correlationalStudy), and hypothesized [mediator](#sec-mediation).
In this case, we predict that receiving touchdowns partially accounts for the association between Wide Receiver's target share and their fantasy points.
This is a silly example because fantasy points are derived, in part, from touchdowns, so of course touchdowns will partially account for almost any effect on Wide Receivers' fantasy points.
This example is merely for demonstrating the process of developing and examining a mediation model.

```{r}
mediationModel <- '
  # direct effect (cPrime)
  fantasyPoints_transformed ~ direct*target_share

  # mediator
  receiving_tds ~ a*target_share
  fantasyPoints_transformed ~ b*receiving_tds

  # indirect effect = a*b
  indirect := a*b

  # total effect (c)
  total := direct + indirect
'
```

To get a robust estimate of the indirect effect, we obtain bootstrapped estimates from 1,000 bootstrap draws.
Typically, we would obtain bootstrapped estimates from 10,000 bootstrap draws, but this example uses only 1,000 bootstrap draws for a shorter runtime.

```{r}
mediationFit <- sem(
  mediationModel,
  data = player_stats_seasonal,
  se = "bootstrap",
  bootstrap = 1000, # generally use 10,000 bootstrap draws; this example uses 1,000 for speed
  parallel = "multicore", # parallelization for speed: use "multicore" for Mac/Linux; "snow" for PC
  iseed = 52242, # for reproducibility
  missing = "ML",
  estimator = "ML",
  std.lv = TRUE,
  fixed.x = FALSE)
```

```{r}
summary(
  mediationFit,
  fit.measures = TRUE,
  standardized = TRUE,
  rsquare = TRUE)
```

We can also estimate a model with multiple hypothesized [mediator](#sec-mediations):

```{r}
multipleMediatorModel <- '
  # direct effect (cPrime)
  fantasyPoints_transformed ~ direct*target_share

  # mediator
  receiving_tds ~ a1*target_share
  receiving_yards ~ a2*target_share

  fantasyPoints_transformed ~ b1*receiving_tds + b1*receiving_yards

  # indirect effect = a*b
  indirect1 := a1*b1
  indirect2 := a2*b2
  indirectTotal := indirect1 + indirect2

  # total effect (c)
  total := direct + indirectTotal
'
```

```{r}
multipleMediatorFit <- sem(
  multipleMediatorModel,
  data = player_stats_seasonal,
  se = "bootstrap",
  bootstrap = 1000, # generally use 10,000 bootstrap draws; this example uses 1,000 for speed
  parallel = "multicore", # parallelization for speed: use "multicore" for Mac/Linux; "snow" for PC
  iseed = 52242, # for reproducibility
  missing = "ML",
  estimator = "ML",
  std.lv = TRUE,
  fixed.x = FALSE)
```

```{r}
summary(
  multipleMediatorFit,
  fit.measures = TRUE,
  standardized = TRUE,
  rsquare = TRUE)
```
